---
title: "Chapter 6: Logistic Regression for Categorical Outcome Variables"
subtitle: "Statistical Modelling Methods: A Guide for the Social and Behavioural Sciences (2nd ed.)"
author: "David Flora & Miranda Too"
format:
  html:
    self-contained: true
    toc: true
    toc-float: true
    theme: flatly
    highlight: haddock
    number-sections: false
    code-overflow: wrap
---

``` {r include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
# Chapter 6 Textbook Content

## Loading R Packages

```{r}
library(car)
library(mediation)
library(misty)
library(stats)
library(ggpubr)
```

## Importing Data

Many R packages include data that can be used for demonstration or practice. In this example, we will use the `jobs` data from the `mediation` package. To create a data frame object, you simply call the `data()` function.

```{r}
data("jobs")
```

For the purposes of our research example, we will make some modifications to the `jobs` data. First, we want to create a binary "work" variable to represent whether the participant was employed at the time of the follow-up interview. We want to recode the values such that 0 = unemployed and 1 = employed.

```{r}
#Here we are using an ifelse statement to create a new variable. 

#This function call indicates that the new 'work' variable will equal 0 if a case has been recorded as "psyump" (unemployed) in the original 'work1' variable of the jobs data, whereas the new 'work' variable will equal 1 (employed) for all other cases.
jobs$work <- ifelse(jobs$work1 == "psyump", 0, 1)
```

Next, we can create a new data frame (`jobdat`) only containing the variables that we need from the original 'jobs' data frame.

```{r}
jobdat <- jobs[,c("work", "treat", "age", "marital")]
```

## Research Example

### Descriptive Statistics

#### Table 6.1

When assessing relationships between categorical variables, it is helpful to look at their conditional frequencies. One way to create a conditional frequency table is by using the `crosstab()` function from the `misty` package.

```{r}
#Conditional frequencies of work status by treatment category
crosstab(jobdat[, c("treat", "work")], print="row")
```

```{r}
#Descriptive stats for age
descript(jobdat$age, print=c('m','sd','med','min','max','skew','kurt'))
```

```{r}
#Frequency table of marital status
freq(jobdat$marital)
```

## Binary logistic regression model specification

### Table 6.3 Logistic regression model of work regressed on treat and age

The `glm()` function from the `stats` package can be used to fit logistic regression and other generalized linear models. Like the `lm()` function, a formula representing the model specification is included as the first argument. Following the formula argument, the `family=` argument is used to indicate the error distribution and link function for the model. For a binary logistic regression model, the family is specified as `binomial`.

```{r}
mod1 <- glm(work ~ treat + age, family=binomial, data=jobdat)
summary(mod1)

#The output indicates that the treat effect is not significant (p = .065), but the age effect is significant.
#But as explained in the text, the slope estimates need to be converted to odds ratios for interpretation.
```

The `coef()` and `confint()` functions from the stats package can be wrapped within the `exp()` function to produce exponentiated coefficients (i.e., odds ratios) and confidence intervals. As discussed in Chapter 6, exponentiating aids interpretation by converting the slope estimates into odds ratios.

```{r}
#Exponentiated estimates and confidence intervals
exp(coef(mod1))

#Receiving the treatment multiplies the odds of employment by 1.33.
#The odds of employment is 96.58% lower for each additional year of age.

exp(confint(mod1))

#Because the treat effect is not significant, the CI for its odds ratio crosses 1.
```

As mentioned in Chapter 6, it might be preferable to obtain the effect for a 10-year age difference instead of a 1-year age difference. Thus, we can multiply the linear slope estimate for age by 10, then exponentiate the result.

```{r}
#the linear age estimate is the third coefficient estimate from mod1
exp(10*mod1$coef[3])

#The odds of employment is 70.667% lower per 10 years of age.
```

### Likelihood Ratio Test for Comparing Nested Models

When including a categorical variable (a `Factor`-type variable) in a model, R will automatically select a reference category and represent the K categories with k-1 dummy variables (see the Chapter 3 material for further explanation).

In this example, we are adding the `marital` variable, which has 5 categories. There will be 4 dummy variables, and the reference category will be whichever category is currently specified as the first level of the `marital` factor. You can use the `levels()` function to check the levels of a variable and find out what the reference category is:

```{r}
levels(jobdat$marital) 
#nevmarr is the reference category
```

Here, `nevmarr`, or "Never Married", is the reference category because it is listed first. If you want to change your reference category, you can do so using the `relevel()` function in the `stats` package.

```{r}
#Specify reference category using the ref= argument
jobdat$marital <- relevel(jobdat$marital, ref= "nevmarr")
#Here, we're just demonstrating the function, and leaving "nevmarr" as the reference category.
```

```{r}
#Fitting an expanded model with the marital status dummy variables
mod2 <- glm(work ~ treat + age + marital, family=binomial, data=jobdat)
summary(mod2)

#The output contains a logistic regression slope estimate (and t-test) for each dummy variable. For example:
#"maritalmarried" compares the "married" category with "nevmarr"
#Being married doesn't significantly effect the work outcome (p = .6395), holding treat and age constant.
```

Like other effects in logistic regression, we need to exponentiate the dummy variable slope estimates for interpretation:

```{r}
exp(coef(mod2))

#The odds of a married person being employed is 1.09 times the odds of a never-married person being employed.

#The odds of a widowed person being employed is 74% of the odds of a never-married person being employed.
```

```{r}
#Likelihood ratio test comparing less complex model to more complex model
anova(mod1, mod2, test = "Chisq")

#Because Model 2 does not fit significantly better than Model 1, conclude that marital status is not significantly related to employment, holding treat and age constant.
```

### Logistic Regression Model Diagnostics

#### Figure 6.2 Assumption checks

The `residualPlots()` function from the `car` package is handy for creating diagnostic plots for logistic regression as well as basic linear models.

```{r}
#Plot of Pearson residuals against predicted logists of age using the car package. 

#Here we include all of the terms of the model except for "treat," because it is binary
#We have also specified that a loess line should be included in the plot

residualPlots(mod1, terms = ~. -treat, smooth=list(smoother=loessLine, span=2/3, col=carPalette()[1]))
```

The `influenceIndexPlot()` function from the `car` package can create an index plot of the Cook's distance values.

```{r}
#Cook's distance Index plot. 
influenceIndexPlot(mod1, vars="Cook")
```

## Logistic Regression for Outcomes with More than Two Categories

### Research Example

For this research example, we are using data from the World Values Survey of 1995-1997, included in the `carData` package.

#### Importing Data

```{r}
library(carData)
data(WVS)
summary(WVS)
```

For this example, we won't use the gender variable, so we can remove it as we save the data frame into a new object:

```{r}
#Removing the 6th column and creating wvsdat data frame
wvsdat <- WVS[,-6]
```

```{r}
#Renaming the poverty variable and specifying that it is a factor (i.e., categorical).
wvsdat$attitude <- as.factor(wvsdat$poverty)
```

### Graphs of variable distributions

```{r}
#Boxplot of distribution of age by attitude
boxplot(wvsdat$age ~ wvsdat$attitude, horizontal = T)

#Boxplot of distribution of age by attitude and religion. 

#The "facet.by" argument is used to specify the variable used for creating individual plots.
ggboxplot(wvsdat, y="age", x="attitude", facet.by = "religion", orientation="horizontal",
          palette = c("#00AFBB", "#E7B800"))

```

### Table 6.4: Multinomial model of attitude regressed on age and religion


```{r}
#make 'about right' the reference category.
#To use the relevel() function, the attitude variable first needs to be an un-ordered factor:
wvsdat$attitude2 <- factor(wvsdat$attitude, ordered=F)
wvsdat$attitude2 <- relevel(wvsdat$attitude2, ref = "About Right")
```

To estimate a multinomial model, we can use the `multinom()` function in the nnet package. Like the `lm()` and `glm()` functions, it uses the same formula syntax to represent the model parameters.

```{r}
library(nnet)
#multinomial model - attitude regressed on religion and age
mnom1 <- multinom(attitude2 ~ religion+age, data = wvsdat)

summary(mnom1, Wald = T)

#None of this output is particularly interpretable because we will need to exponentiate the coefficient estimates (see below), but the Wald stats are Z-test statistics Wald stats for age are big, Wald stats for religion are small.
```

We can use the `pnorm()` function to get two-tailed p-values from the Wald Z-statistics:

```{r}
#getting two-tailed p-values for age using Wald statistics
pnorm(4.395226, 0, 1, lower.tail = F)*2
#age is significantly associated with the probability of saying "too little" instead of "about right"
pnorm(5.419484, 0, 1, lower.tail = F)*2
#age is significantly associated with the probability of saying "too much" instead of "about right"
```

```{r}
#getting two-tailed p-values for religion using Wald statistics
pnorm(0.4383507, 0, 1, lower.tail = F)*2
pnorm(0.6629923, 0, 1, lower.tail = F)*2

#the religion effects are not significant.
```

Like binary logistic regression, we can use the `exp()` function paired with the `coef()` and `confint()` functions to convert slope estimates to odds ratios and get CIs for the odds ratios.

```{r}
#exponentiating estimates to obtain odds ratios 
exp(coef(mnom1))
#The odds of saying "too little" instead of "about right" is 99.22% lower for each additional year of age.

#The odds of saying "too much" instead of "about right" is 1.01323 times greater for each additional year of age.

exp(confint(mnom1))

#Consistent with the Wald tests, the CIs for religion cross 1 (non-significant) whereas the CIs for age do not cross 1 (significant).
```

The text mentions that the age or religion effects might vary by country. That is, country could moderate those effects. Here, we apply techniques introduced in Chapter 4 to test and probe the interaction between country and religion:

```{r}
#By default, country (a factor variable) will be dummy coded, with Australia as the reference category

mnom2a <- multinom(attitude2 ~ religion+age+country, data = wvsdat)

#Expanded model with country and its interaction with religion.
mnom2b <- multinom(attitude2 ~ religion+age+country+(religion*country), data = wvsdat)

#Model comparison with likelihood ratio test
anova(mnom2a ,mnom2b)

#The likelihood ratio test indicates that country significantly interacts with religion. 
```

To begin probing the interaction, let's get odds ratios (and their CIs) for the interaction model:

```{r} 
#Because Australia is the reference category for country, the lower-order effects of religion pertain to Australia:
exp(coef(mnom2b))
exp(confint(mnom2b))
#The CI around the religion odds ratio crosses 1 for the comparison of "too little" with "about right" (CI: 0.97, 1.59)
#Therefore, among Australian respondents, religion is not significantly associated with the odds of saying "too little" instead of "about right".

#But the CI around the religion odds ratio does NOT cross 1 for the comparison of "too much" with "about right" (CI: 1.159, 2.45)
#Therefore, among Australian respondents, religion is significantly associated with the odds of saying "too much" instead of "about right".
#The religion odds ratio for "too much" (1.686) indicates that Australian respondents who are a member of a religion have an odds of responding "too much" instead of "about right" that is 1.686 times greater than the odds for Australian respondents who are not a member of a religion.
```

To continue probing the interaction, we can change the reference category and re-estimate the model. Here, we change the country reference category to USA so that the lower-order religion effects pertain to the USA:

```{r}
#Use the relevel() function to change the reference category
wvsdat$country <- relevel(wvsdat$country, ref= "USA")

#Then re-estimate the interaction model:
mnom2c <- multinom(attitude2 ~ religion+age+country+(religion*country), data = wvsdat)
#get odds ratios and their CIs:
exp(coef(mnom2c))
exp(confint(mnom2c))

#The CI around the religion odds ratio does not cross 1 for the comparison of "too little" with "about right" (CI: 0.48, 0.92)
#Therefore, among USA respondents, religion is significantly associated with the odds of saying "too little" instead of "about right".
#The religion odds ratio for "too much" (0.669) indicates that USA respondents who are a member of a religion have an odds of responding "too little" instead of "about right" that is 66.9% of the odds for USA respondents who are not a member of a religion.

#The CI around the religion odds ratio crosses 1 for the comparison of "too much" with "about right" (CI: 0.834, 1.728).
#Therefore, among USA respondents, religion is not significantly associated with the odds of saying "too much" instead of "about right".
```

We could proceed in the same manner to get religion effects for Norway and Sweden (i.e., change the reference category, re-estimate the model, interpret the lower-order religion effects as specific to the refrence category country). Readers are invited to do so on their own.

## Ordinal Logistic Regression

A proportional odds ordinal logistic regression model can be estimated using the `polr()` function in the `MASS` package.

```{r}
#Notice that we are using the original 'poverty' variable as the outcome.
#Check that 'poverty' is an ordinal factor, with "too little" as the lowest category:
str(wvsdat$poverty)

library(MASS)
#Fit the ordinal regression 
ordmod <- polr(attitude ~ age + religion, data = wvsdat)
```

To compare the multinomial model fit to the ordinal logistic model fit, we can obtain the residual deviance values for each model using the `deviance()` function, and save them into objects. We can then subtract the residual deviance of the less complex model (the multinomial model) from the more complex model (the ordinal/proportional odds model) to obtain the likelihood ratio test statistic. We can then use the `pchisq()` function to compare this value against a chi-square distribution with a specified number of degrees of freedom (in this case, it is 2).

```{r}
#Save residual deviances into respective objects
mod1.d <- deviance(mnom1) 
mod2.d <- deviance(ordmod)

#Obtain difference in residual deviance (i.e., likelihood ratio )
LR <- mod2.d - mod1.d

# printing the likehihood ratio test statistic, reported as chi-square = 7.92 in the text.
LR 

pchisq(LR, 2, lower.tail = F)
#Rounds to .02, as presented in the text.
```

```{r}
#Additionally, we can obtain the AIC values for each model by using the AIC() function
AIC(mnom1)
AIC(ordmod)
```

### Table 6.5: Ordinal Model Summary

```{r}
#Model summary 
summary(ordmod)

#Although the output doesn't include p-values, the t-statistic for age is so large that it's p-value must be p < .001.
#For religion, the t-stat (-0.9994) is rather weak, but we can get the exact p-value with
pnorm(0.9994, 0, 1, lower.tail = F)*2

#As with other logistic regression models, the estimates need to be exponentiated (i.e., converted to odds ratios) for interpretation.

exp(coef(ordmod))
exp(confint(ordmod))

#As reported in the text, the OR for age indicates that a one-year age difference multiplies the odds of a response in a lower category by 1.01 (recall the negative sign in Equation 6.8). 
```
