---
title: "Chapter 3: Multiple Regression with Categorical Regressors"
subtitle: "Statistical Modelling Methods: A Guide for the Social and Behavioural Sciences (2nd ed.)"
author: "David Flora & Miranda Too"
format:
  html:
    self-contained: true
    toc: true
    toc-float: true
    theme: flatly
    highlight: haddock
    number-sections: false
    code-overflow: wrap
---

# Chapter 3 Textbook Content

## Loading R Packages

```{r}
library(misty)
```

## Research Example

This artificial data provided by Darlington and Hayes (2017) examines the association between marital status and general life satisfaction.

## Importing data

```{r}
married <- read.csv("married.txt", sep="")
str(married)
```

The marriage variable `mstatus` is coded into four categories: 1 = married, 2 = divorced, 3 = single, and 4 = widowed. Life satisfaction, `satis`, is a continuous variable with possible scores ranging from 0 to 100.

```{r}
#Replacing numerical codes with category labels
married$mstatus <- factor(married$mstatus, labels=c('married','divorced','single','widowed'))
str(married$mstatus)
```

### Table 3.1: Descriptive statistics of life satisfaction by marital status

```{r}
#Life satisfaction descriptive statistics by marital status group 

descript(married$satis, group = married$mstatus, print = c("n","m",'sd','med','skew','kurt'))
```

### Figure 3.1: Boxplot by Outcome Group

Grouped boxplots can be created using the `boxplot()` function. Instead of using a single vector, a formula in the `y ~ x` form can be used, where `x` is a numeric variable, and `y` is a categorical variable.

```{r}
boxplot(married$satis~married$mstatus, ylab='Life satisfaction score', xlab='Marital status')
```

## Dummy-Variable Regression and ANOVA

First, we can create dummy variables in accordance with Table 3.2:

```{r}
#This code creates dummy variables d1, d2, and d3, and puts them in the 'married' data frame.

married$d1 <- ifelse(married$mstatus=="divorced", 1, 0)
#If mstatus is divorced, then d1 = 1, else d1 = 0

married$d2 <- ifelse(married$mstatus=="single", 1, 0)
#If mstatus is divorced, then d2 = 1, else d2 = 0

married$d3 <- ifelse(married$mstatus=="widowed", 1, 0)
#If mstatus is widowed, then d3 = 1, else d3 = 0
```

Then, we can use the dummy variables to represent the nominal marital status variable and fit the model represented in Equation 3.1 to obtain the results in Table 3.3:

```{r}
#Life Satisfaction regressed on marital status
dummyMod <- lm(satis ~ d1+d2+d3, data=married)
summary(dummyMod)

#Notice that the results match the values reported in Table 3.3.

#Get confidence intervals too
confint(dummyMod)
```

Instead of literally creating dummy variables and then using them in the regression model, we can simply refer to the original, factor-type, multi-category variable (`mstatus`) in the model formula, and the 'lm()' function will automatically (by default) treat the factor-variable as a set of dummy variables to fit the model, as shown next:

```{r}

#You can use the `levels()` function to check the levels of a variable and find out what the reference category is:
levels(married$mstatus)
#The first listed category ("married") will be the reference category.

#Life Satisfaction regressed on marital status
anovaMod <- lm(satis ~ mstatus, data=married)
summary(anovaMod)
#Obtain confidence intervals around model parameters
confint(anovaMod)

#Notice that the estimates and confidence intervals are exactly the same as when the model was fitted to the dummy variables!
#Again, each estimate pertains to the comparison of one category with the reference category.
#For example, output for "mstatusdivorced" pertains to a comparison of the divorced category with the married (reference) category.
```

## The ANCOVA Model and Beyond

### Table 3.4

To investigate whether satisfaction scores significantly differ across marital status groups while holding age constant, we can use a hierarchical regression strategy: (1) fit a model with satisfaction regressed on age, then (2) fit a model with satisfaction regressed on age and marital status, and then (3) use an F-test to determine whether R-squared significantly increased.

```{r}
# (1) Fit model with satisfaction regressed on age 
mod0 <- lm(satis ~ age, data=married)
summary(mod0)
#From this output, we mainly just want to note that R^2 = .3309


# (2) Fit a model with satisfaction regressed on martial status and age
ancovaMod <- lm(satis ~ mstatus+age, data=married)
summary(ancovaMod)
#Recall that since "married" was the reference category, each of the estimates for the marital status categories refers to a comparison with those who are married (holding age constant in the `ancovaMod` model.
#But at this point, the output of interest is R^2 = .4084 


# (3) Compare the models with an F-test using the anova function
anova(mod0, ancovaMod)
#This F-test tells us that the R^2 increase from .3309 to .4084 is not significant, p = .2748.
```
### Comparing adjusted means across groups

The `emmeans()` function of the `emmeans` package can calculate adjusted means from an ANCOVA model automatically ("emmeans" stands for "expected marginal means"). Recall from the text that adjusted means are the predicted values of the outcome variable for each level of a categorical independent variable, with a covariate held constant at its mean value.
The first argument of the `emmeans()` function is the name of the ANCOVA model as fitted with `lm()`, and the second argument is \~ followed by the name of the factor-type, categorical independent variable.

```{r}
library(emmeans)
emmeans(ancovaMod, ~mstatus)

#The values listed under the "emmean" column correspond to the adjusted means reported in the text.
```

## Regression Diagnostics with Categorical Regressors

### Figure 3.2: Scatterplot of Studentized Residuals by Age

```{r}
library(car)
scatterplot(rstudent(ancovaMod)~married$age, ylab='Studentized residuals', 
            xlab='Age', boxplot=F, col="black", smooth=list(var=F, span=1.0))

```

### Figure 3.3: Boxplot of Studentized Residuals by Marital Status

```{r}
boxplot(rstudent(ancovaMod)~married$mstatus, ylab='Studentized residuals', xlab='marital status')
```

### Figure 3.4: Q-Q Plot of Model

```{r}
qqPlot(ancovaMod, envelope = F, ylab = "Studentized residuals", grid=F, id=F)
```

## Using Contrast-Code Variables

The following material accompanies the supplementary text for this chapter.

### Creating Helmert contrasts

As described above, by default R will represent a factor-type variable using a set of dummy variables.
We can use the `contrasts()` function to see that the `mstatus` variable is currently represented as a set of three dummy variables with "married" as the reference category:

```{r}
contrasts(married$mstatus)

#Notice how this table corresponds to the table of dummy variable values in Table 3.2 of the text.
```

But we can specify an alternative set of contrasts using the `matrix()` function.
Here, we create a matrix called `helmerts` which contains the contrast-code values given in Table S3.1 in the supplementary text. Because we need to create three contrasts, the matrix has three columns (`ncol=3`). 
Then, we assign this new `helmerts` matrix to the contrasts representing `mstatus`, thereby overriding the dummy variables:

```{r}
helmerts <- matrix(c(-1/3,-1/3,  1,-1/3,
                        1,-1/2,  0,-1/2,
                        0,   1,  0,-1),
                      ncol=3)
contrasts(married$mstatus) <- helmerts

#check what the contrasts look like now:
contrasts(married$mstatus)

#The first column corresponds to the "C1" variable in the supplementary text (single vs. others)
#The 2nd column corresponds to the "C2" variable (married vs. divorced and widowed)
#The 3rd column corresponds to the "C3" variable (divorced vs. widowed)
```

Now, when we regress the `satis` outcome on `mstatus`, we get results based on these contrasts instead of dummy variables:

```{r}
mod.helm <- lm(satis ~ mstatus, data=married)
summary(mod.helm)
confint(mod.helm)

#Notice how the results match those in supplementary Table S3.2
```

### Improved contrasts

Next, we can change the contrasts to match the 'improved contrasts' represented in Table S3.3:

```{r}
helmert2 <- matrix(c(-1/4,-1/4,3/4,-1/4,
                      2/3,-1/3,  0,-1/3,
                      0,    .5,  0,-.5),
                      ncol=3)
contrasts(married$mstatus) <- helmert2

#check what the contrasts look like now:
contrasts(married$mstatus)
```

And then fit the model again:

```{r}
mod.helm <- lm(satis ~ mstatus, data=married)
summary(mod.helm)
confint(mod.helm)

#Now the results match those in supplementary Table S3.4
```
